\begin{otherlanguage}{english}
\begin{abstract}
Many real-world problem settings give rise to \NP-hard combinatorial optimization problems.
This results in a need for non-trivial algorithmic solutions for finding optimal solutions to such problems.
Many such solutions---ranging from probabilistic and meta-heuristic algorithms to declarative programming---have been presented for optimization problems with a single objective.
Less work has been done on approaches to optimization problems with \emph{multiple} objectives.

We present \algname{}, an exact declarative approach for finding so-called Pareto-optimal solutions to bi-objective optimization problems.
A bi-objective optimization problem arises for example when learning interpretable classifiers and the size as well as the classification error of the classifier should be taken into account as objectives.
Using propositional logic as the declarative programming language, we seek to extend the progress and success in maximum satisfiability (MaxSAT) solving to two objectives.
The approach can be viewed as an instantiation of the lexicographic method and makes use of a single SAT solver that is preserved throughout the entire search procedure.
\algname{} allows for solving three tasks for bi-objective optimization:
finding a single Pareto-optimal solution, finding one representative solution for each Pareto point, and enumerating all Pareto-optimal solutions.

We provide an open-source implementation of five variants of \algname{}, building on different algorithms proposed for MaxSAT.
We empirically evaluate these five variants, comparing their runtime performance to that of three competing algorithmic approaches.
The empirical comparison shows practical benefits of our approach in the contexts of learning interpretable decision rules and bi-objective set covering.
Furthermore, for the best-performing variant of \algname{}, we study the effects of proposed refinements to determine their effectiveness.
\end{abstract}
\end{otherlanguage}
