\appendix{Datasets Used for Decision Rule Learning\label{appendix:datasets}}

\Cref{tab:datasets} summarizes the datasets used in the empirical evaluations, including their origin and statistics, as well as the sizes of CNF formulas obtained from them with the encoding from~\textcite{DBLP:conf/cp/MaliotovM18}.
The original files were downloaded from the UCI Machine Learning Repository~\autocite{UciMlr} and from Kaggle ({\small\url{https://www.kaggle.com}}).
We randomly and independently sampled subsets of $\nsamp\in\{50,100,1000,5000,10000\}$ data samples from the datasets, four of each size (when applicable), resulting in a         
total of 372 datasets, and discretized the data as in~\autocite{DBLP:conf/cp/MaliotovM18}:
categorical features are one-hot encoded, continuous features discretized by comparing to a collection of thresholds. 

In addition to the name and the source of the datasets, the table shows the number of data samples as well as the number of features before and after discretization.
The last two columns give some statistics about the formulas generated with the encoding~\autocite{DBLP:conf/cp/MaliotovM18} for two clauses based on the full datasets.
We report both the number of clauses and the number of variables in these formulas.

For the decision rule instances, the instance that took the longest time to solve that did not time out for the \msh{} variant was a subset of 100 samples of the Connect 4 dataset.
The formula of this dataset has 678 variables and 4152 clauses.
The largest instance in terms of the number of samples that our algorithm was able to find a representative for every Pareto-point for was a subset of the Travel Insurance dataset with 10000 samples.
When looking at the number of features, the largest solvable dataset was a subset of the Twitter dataset with 50 samples and 1511 discretized features.

\begin{sidewaystable}
    \centering
    \caption{The datasets used in the decision rule experiments and some summary statistics about them and the encoded formulas created from them.}\label{tab:datasets}
    {\small
    \begin{tabular}{@{}llS[table-format=6.0]S[table-format=3.0]S[table-format=4.0]S[table-format=5.2]S[table-format=3.3]@{}}
        \toprule
        Dataset & Source & {\# samples} & {\# features} & {\# disc. feat.} & {\# clauses ($10^3$)} & {\# vars ($10^3$)} \\
        \midrule
        Adult                                & UCI    &  32561 &  14 &  144 &   635    &   98.1 \\
        Bank Marketing                       & UCI    &  45211 &  16 &   88 &  1329    &  136 \\
        Banknote Authentication              & UCI    &    372 &   4 &   16 &     6.67 &    4.16 \\
        Connect 4                            & UCI    &  67557 &  42 &  126 &  2052    &  203 \\
        Default of Credit Card Clients       & UCI    &  30000 &  23 &  110 &   878    &   90.3 \\
        Dota 2 Games Results                 & UCI    &  92650 & 115 &  345 & 11164    &  279 \\
        FIFA 2018 Man of the Match           & Kaggle &    128 &  26 &  106 &     3.00 &    0.708 \\
        Heart Disease                        & Kaggle &    303 &  13 &   31 &     3.72 &    1.00 \\
        Indian Liver Patient Dataset         & UCI    &    583 &  10 &   14 &     6.67 &    1.79 \\
        Ionosphere                           & UCI    &    351 &  33 &  144 &     9.90 &    1.49 \\
        Iris                                 & UCI    &    150 &   4 &   11 &     1.08 &    0.483 \\
        MAGIC Gamma Telescope                & UCI    &  19020 &  10 &   79 &   273    &   57.3 \\
        Medical Hospital Readmissions        & Kaggle &  25000 &  64 &  125 &  1641    &   75.4 \\
        Mushroom                             & UCI    &   8124 &  22 &  115 &   190    &   24.7 \\
        Parkinsons                           & UCI    &    195 &  22 &   51 &     2.81 &    0.738 \\
        Pima Indians Diabetes                & Kaggle &    768 &   8 &   30 &     7.25 &    2.39 \\
        Skin Segmentation                    & UCI    & 245057 &   3 &  119 &   745    &  736 \\
        Tic-Tac-Toe Endgame                  & UCI    &    958 &   9 &   27 &     7.75 &    2.96 \\
        Buzz in Social Media (Toms Hardware) & UCI    &  28179 &  96 &  910 &  3712    &   87.3 \\
        Buzz in Social Media (Twitter)       & UCI    &  49999 &  77 & 1511 &  5406    &  155 \\
        Blood Transfusion Service Center     & UCI    &    748 &   4 &    6 &     4.39 &    2.26 \\
        Travel Insurance                     & Kaggle &  63326 &  10 &  211 &  1188    &  191 \\
        Wisconsin Diagnostic Breast Cancer   & UCI    &    569 &  30 &   88 &    20.7  &    1.97 \\
        Rain in Australia                    & Kaggle & 107696 &  16 &  141 &  2952    &  339 \\
        \bottomrule
    \end{tabular}
    }
\end{sidewaystable}